Index-aware learning of circuits

### 2024-04-05 doc/article/arxiv/article_highlighted.tex
graphical abstract
final revision

\textcolor{red}{In order to obtain a version of \eqref{eq:md_mnai} that is better suited to analysis and implementation, we consider the device function Jacobians
\begin{align}
    \mb{G} \big( \AT{R} \vphi \big) \coloneqq \mb{J}_{\AT{R} \vphi}(\gR), \quad \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC), \label{eq:md_dfj}
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}.} Inserting \eqref{eq:md_dfj} into the original system, while linearizing $\gR$ and writing everything in matrix form, yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0}, \label{eq:md_mnaii}
\end{align}
assuming that $\phiL$ and $\qC$ do not explicitly depend on time.

\textcolor{red}{We note that \eqref{eq:md_mnaii} readily extends to multiport devices\cite{tischendorf1999}. A general inductive $n$-port for example, can also be modeled by a function $\phiL(\i{L}) = \big[ \phi_1(\i{L}), \dotsc, \phi_{n-1}(\i{L}) \big]^\T$, however the Jacobian is not necessarily diagonal in this case as the component functions of $\phiL$ can each depend on all the currents $\i{L} = [i_1, \cdots, i_{n-1}]^\T$. Regarding the incidence matrices, one chooses a reference node $\varphi_0$ for the multiport device and then considers $n-1$ branches from the remaining $n-1$ nodes to the reference. Interpreting the multiport as a single node and using Kirchhoff's current law then gives $i_0 = -\sum_{k=1}^{n-1} i_k$ for the reference current, when orienting all currents to point toward the device. Following this approach, there is then no difference in treating multiports compared to one-ports\cite{tischendorf1999}.}

\subsection{Index concepts}
\label{subsec:ic}
\textcolor{red}{Before outlining the dissection index, we want to give a brief introduction to index concepts more generally. There are multiple definitions of the index of a DAE, along with related index concepts that each possess different strengths and weaknesses\cite{mehrmann2015}. One important aspect that unites these ideas is that they agree in key cases, e.g.~when looking at linear DAEs, and the same also holds true for the dissection index. To emphasize the practical importance of the notion of index, we take a closer look at the perturbation index. It is based on a perturbed version of the DAE \eqref{eq:int_dae} (we leave out the parameters $\mb{p}$ for conciseness)
\begin{align}
    \mb{F}(\mbh{x}', \mbh{x}, t) = \boldsymbol{\varepsilon}(t), \quad \mbh{x}(0) = \mbh{x}_0, \label{eq:ic_pdae}
\end{align}
where $\varepsilon$ is a sufficiently smooth perturbation, such that the required derivatives exist.
\begin{definition}
    \label{def:ic_pi}
    Let $\mb{x}$ be a solution of the unperturbed DAE, then the DAE is said to have perturbation index $\nu \in \mathbb{N}$, if $\nu$ is the smallest natural number such that, for any sufficiently smooth solution $\mbh{x}$ of \eqref{eq:ic_pdae}, there exists $c \in \mathbb{R}$ with
    \begin{align*}
        \| \mbh{x} - \mb{x} \| \leq c \big( \| \mbh{x}_0 - \mb{x}_0 \| + \| \boldsymbol{\varepsilon} \|_\infty + \| \boldsymbol{\varepsilon}' \|_\infty + \dotsb + \| \boldsymbol{\varepsilon}^{(\nu - 1)} \|_\infty \big)
    \end{align*}
    for an appropriate norm $\| \cdot \|$ and the right hand side small enough.
\end{definition}
The idea behind this definition is to capture the impact of perturbations on the solution, as the name suggests. Usually these pertubations are assumed small, in the sense that $\| \boldsymbol{\varepsilon} \|_\infty \ll 1$, but fast changing, such that $\| \boldsymbol{\varepsilon}^{(\nu)} \|_\infty$ may grow very quickly in $\nu$.}

\paragraph{Example}
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=0.35\textwidth]{vcr}
    \end{center}
    \caption{Small example circuit for illustrating the perturbation index.}
    \label{fig:ic_vcr}
\end{figure}
\textcolor{red}{To illustrate the perturbation index, and also to hint at its relevance for circuit simulation, we consider the small example given in \autoref{fig:ic_vcr}. Applying MNA to the circuit and introducing a perturbation yields the following DAE
\begin{subequations}
    \label{eq:ic_vcr}
    \begin{align}
        C \ddt \hat{\varphi}_1 + \frac{1}{R} \hat{\varphi}_1 + \hat{i}_\mr{V} &= \varepsilon_1(t) \label{eq:ic_vcra}\\
        -\hat{\varphi}_1 + v_\mr{s}(t) &= \varepsilon_2(t), \label{eq:ic_vcrb}
    \end{align}
\end{subequations}
where $\mbh{x} = [\hat{\varphi}_1, \hat{i}_\mr{V}]^\T$. Using \eqref{eq:ic_vcrb} we find
\begin{align}
    \hat{\varphi}_1 = v_\mr{s}(t) - \varepsilon_2(t). \label{eq:ic_vcrphi}
\end{align}
Inserting \eqref{eq:ic_vcrphi} into \eqref{eq:ic_vcra} and rearranging then gives
\begin{align*}
    \hat{i}_\mr{V} &= \varepsilon_1(t) - C \ddt \big( v_\mr{s}(t) - \varepsilon_2(t) \big) - \frac{1}{R} \big( v_\mr{s}(t) - \varepsilon_2(t) \big).
\end{align*}
Noting that the solution $\mb{x}$ to the unperturbed problem follows directly from the perturbed solution by setting $\boldsymbol{\varepsilon} = \mb{0}$, we obtain
\begin{align}
    \| \mbh{x} - \mb{x} \| = \begin{Vmatrix}
        -\varepsilon_2(t)\\
        \varepsilon_1(t) + C \ddt \varepsilon_2(t) + \frac{1}{R} \varepsilon_2(t)
    \end{Vmatrix}, \label{eq:ic_vcrpi}
\end{align}
so the unperturbed system corresponding to \eqref{eq:ic_vcr} has perturbation index $\nu = 2$ as \eqref{eq:ic_vcrpi} depends on the first derivative of $\varepsilon_2$.}

@@

We note that all basis functions of the second step are also constant for this example, as was the case for the first step. As hinted at in \autoref{sec:di}, it is possible to find a purely topological decoupling based on the dissection index\cite{jansen2014}, which agrees with the intuition given by the index result from \autoref{th:ic_imna}. \textcolor{red}{When considering large circuits, this topological decoupling along with its topological basis functions is to be preferred over other basis function choices, as it avoids the numerical computation of the basis functions, which becomes prohibitively expensive for large systems.} We also note that a similar topological decoupling result exists for circuits including controlled sources, however only for the first step of the decoupling, as it is framed in the context of semi-explicit methods for which one only requires a DAE in semi-explicit form\cite{jansen2014}. An extension of this result to circuits of higher index is within the scope of further research.

@@

\paragraph{Index one case}
\textcolor{red}{In order to recover the algebraic variables at time $t$ in the index one case, we only need to solve \eqref{eq:ioc_iob} for $\mbb{x}(t)$ using the learned $\mbt{x}(t)$.}

\paragraph{Index two case}
\textcolor{red}{In the index two case, we start by solving \eqref{eq:itc_xtp} for $\mbt{x}_\mb{P}(t)$ using the learned $\mbt{x}_\mb{Q}(t)$. Since we also require the derivative $\ddt \mbt{x}_\mb{P}(t)$ in \eqref{eq:itc_itiiib}, we consider a small time increment $\Delta t$ and approximate the derivative using a backward difference
\begin{align*}
    \ddt \mbt{x}_\mb{P}(t) \approx \frac{\mbt{x}_\mb{P}(t + \Delta t) - \mbt{x}_\mb{P}(t)}{\Delta t}.
\end{align*}
We note that this only reflects our implementation; in principle any finite difference (or similar) approximation is possible. Finally, we determine $\mbb{x}_\mb{P}(t)$ and $\mbb{x}_\mb{Q}(t)$ by jointly solving \eqref{eq:itc_xbp} and \eqref{eq:itc_itiiib} using the learned $\mbt{x}_\mb{Q}(t)$, $\mbt{x}_\mb{P}(t)$ and the approximation of $\ddt \mbt{x}_\mb{P}(t)$.}

@@

There is yet another, maybe less expected, benefit that might occur. Looking at the decoupled system from \eqref{eq:itc_do2} we find that the resistance parameter $R$ and the inductance parameter $L$ only appear in the algebraic equation. In terms of our original goal of speeding up design optimization or uncertainty quantification, where a lot of solutions for varying parameter values are required, this means the following: instead of having to solve the full system for a given combination of $R$ and $L$, we can instead simply solve the algebraic equation to obtain the full solution. While the algebraic equation might be more complicated than the simple linear relations of \eqref{eq:itc_do1} and \eqref{eq:itc_do2} in general, solving it is almost certainly much faster than having to integrate the entire system in time. This becomes an even bigger advantage when knowledge about the solution is only required for specific points in time, since the algebraic equation may be solved pointwise. \textcolor{red}{As of now, we have no easy way to automatically determine which parameters appear in the ODE. But when combined with a sequential learning strategy, such as the one outlined in \autoref{sec:ne}, there may still be computational savings due to the learning method requiring less samples for the parameters not appearing in the ODE.}

@@

\paragraph{First example circuit}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do1_solution_phi} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{do1_solution_i_L}
    \end{center}
    \caption{Solution of the first example circuit for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$.}
    \label{fig:ne_do1_solution}
\end{figure}
We again consider the example of \autoref{fig:itc_do1}, where we choose $v_\mr{s}(t) = \sin(600 \pi t)\, \si{\volt}$, $R = \SI{500}{\ohm}$ and the diode is modeled by $g_\mr{D}(\varphi_3) = 10^{-14} \big( e^{(\varphi_3 / \SI{26}{\mV})} - 1 \big)\, \si{\siemens}$. The remaining parameter values are chosen according to the sequential sample selection strategy with $\SI{1}{\milli\henry} \leq L \leq \SI{3}{\milli\henry}$ and $\SI{100}{\nano\farad} \leq C \leq \SI{300}{\nano\farad}$, i.e.~$\mb{p} = [L, C]^\T$ in the context of \autoref{sec:int}. The starting points for the strategy are given by all combinations of the boundary values for $L$ and $C$ together with $t = \SI{0}{\ms}$ and $t = \SI{10}{\ms}$ once for each combination. \textcolor{red}{We note that the choice to consider $L$ and $C$, and not for example $R$, as parameters here is arbitrary and only serves to illustrate the approach. One could still follow the same approach when considering e.g.~the initial conditions, or the diode model, as being parameterized.} Recalling \eqref{eq:itc_do1}, we see that $\varphi_3$ and $i_\mr{L}$ have to be learned as the differential variables. In addition to these two, we also consider the algebraic variable $\varphi_2$ and we then compare the accuracy of learning $\varphi_2$ directly to recovering it from the two differential variables. \autoref{fig:ne_do1_solution} shows the solution for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$, a combination which is not part of the training data.
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do1_convergence}
    \end{center}
    \caption{Convergence of the relative prediction error $e$, for the variables shown in \autoref{fig:ne_do1_solution}. The final accuracies correspond to approximately 180 different combinations of $L$ and $C$ for $\varphi_3$, 39 combinations for $\varphi_2$ and 37 for $i_\mr{L}$.}
    \label{fig:ne_do1_convergence}
\end{figure}

@@

At this point we also return to the discussion from \autoref{sec:ial} about parameters only appearing in the algebraic equation. During the learning process, the sample selection strategy requested 21 unique values for $C$, all of which required full simulations according to \eqref{eq:itc_do2}. The 7 unique values for $L$, that were requested for learning $\varphi_2$, did not require full simulations however, but rather could be reconstructed from \eqref{eq:itc_do2}. \textcolor{red}{This results from the fact that the ODE in \eqref{eq:itc_do2} does not depend on $L$, such that the solution of $\varphi_3$ also does not depend on that parameter. The same goes for the algebraic variable $i_\mr{L}$, thus $\varphi_2$ and $\varphi_1$ can be reconstructed only based on the knowledge of $\varphi_3$.} We again emphasize that time is also included as a parameter within the sample selection strategy, such that the reconstructions of the algebraic variables only need to be evaluated at the particular points in time that are requested by the strategy, rather than at all time points of the solution.
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do2_error_phi_3} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{do2_error_phi_2}
    \end{center}
    \caption{Differences between the mean prediction and simulation for $\varphi_3$ (left) and between the mean prediction, reconstruction and simulation for $\varphi_2$ (right). The predictions are again made for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$ and correspond to the smallest relative errors from \autoref{fig:ne_do2_solution_convergence}.}
    \label{fig:ne_do2_error}
\end{figure}

@@

\begin{figure}[b]
    \begin{center}
        \includegraphics[width=.4\textwidth]{fwr2_solution_v34} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{fwr2_solution_i_L_1}
    \end{center}
    \caption{Solutions of the differential variables of the full wave rectifier circuit for $T_1 = 65\, \si{\celsius}$ and $T_2 = 85\, \si{\celsius}$.}
    \label{fig:ne_fwr2_solution_d}
\end{figure}
\textcolor{red}{As a third and larger example we consider the rectifier circuit from \autoref{fig:ne_fwr2}. Aside from requiring the full index two implementation from \autoref{sec:ial}, this example also showcases another potential application in the context of electrothermal simulations. Electrothermal simulations are used to investigate the thermal behavior of a circuit, by coupling the power that is dissipated in the circuit to a set of equations describing the temperature distribution in the circuit, and by using the temperatures of some components as parameters for certain device functions. In our case, we again model the diodes as nonlinear resistors, however this time the model also includes a temperature dependence\cite{tietze2008}
\begin{align*}
    g_\mr{D}(v_\mr{D}, T_\mr{D}) = I_\mr{s}(T_\mr{D}) \frac{qk}{T_\mr{D}} e^{(qk/T_\mr{D}) v_\mr{D}},
\end{align*}
where $v_\mr{D}$ is the voltage across the diode in forward direction, $T_\mr{D}$ is the temperature of the diode, $q$ is the charge of an electron, $k$ the Boltzmann constant and $I_\mr{s}(T_\mr{D})$ the temperature dependent reverse saturation current\cite{tietze2008}. The current source is given by $i_\mr{s}(t) = 0.1 \cos(100 \pi t)\, \si{\ampere}$ and the capacitor and resistor are linear with $C = 1\, \si{\milli\farad}$, $R = 50\, \si{\ohm}$. The circuit also contains a transformer $\mr{T}_{12}$ modeled as a nonlinear inductive multiport with
\begin{align*}
    \mb{L}(\i{L}) = \begin{bmatrix}
        L(i_{\mr{L}_1}) & \sqrt{0.09 L(i_{\mr{L}_1}) L(i_{\mr{L}_2})}\\
        \sqrt{0.09 L(i_{\mr{L}_1}) L(i_{\mr{L}_2})} & 0.1 L(i_{\mr{L}_2})
    \end{bmatrix},
\end{align*}
where $L(i_\mr{L})$ is a fourth order polynomial in $i_\mr{L}$\cite{lullo2017}. We also note that rectifiers play an important role in many power electronics applications; they are key when converting high AC to lower DC voltages. In the example we still use a current source, since this leads to an index two circuit in this case, compare \autoref{th:ic_imna}, which requires the full index two approach from \autoref{sec:ial}.}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{fwr2_solution_phi} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{fwr2_solution_i_L_2}
    \end{center}
    \caption{Solutions of the algebraic variables of the full wave rectifier circuit for $T_1 = 65\, \si{\celsius}$ and $T_2 = 85\, \si{\celsius}$.}
    \label{fig:ne_fwr2_solution_a}
\end{figure}

>>>>

\textcolor{red}{In order to obtain a version of \eqref{eq:md_mnai} that is better suited to analysis and implementation, we consider the device function Jacobians\footnote{\textcolor{red}{Note that the definition of $\mb{G}$ via the Jacobian only serves to obtain a matrix form in \eqref{eq:md_mnaii} and has no impact on the general approach. As such, one may also work directly with the original system from \eqref{eq:md_mnai}, however software implementations of MNA often work with $\mb{G}$ as defined in \eqref{eq:md_dfj} (e.g.~when using Newton's method for solving nonlinear systems).}}}
\begin{align}
    \mb{G} \big( \AT{R} \vphi \big) \coloneqq \mb{J}_{\AT{R} \vphi}(\gR), \quad \mb{L}(\i{L}) \coloneqq \mb{J}_{\i{L}}(\phiL), \quad \mb{C} \big( \AT{C} \vphi \big) \coloneqq \mb{J}_{\AT{C} \vphi}(\qC), \label{eq:md_dfj}
\end{align}
where we use the same notation for the Jacobians as in \autoref{sec:int}. Inserting \eqref{eq:md_dfj} into the original system and writing everything in matrix form yields
\begin{align}
    \begin{bmatrix}
        \A{C}^{\phantom{\T}} \mb{C} \big( \AT{C} \vphi \big) \AT{C} & &\\
        & \mb{L}(\i{L}) &\\
        & & \mb{0}
    \end{bmatrix} \ddt \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{R}^{\phantom{\T}} \mb{G} \big( \AT{R} \vphi \big) \AT{R} & \A{L} & \A{V}\\
        -\AT{L} & &\\
        -\AT{V} & &
    \end{bmatrix} \begin{bmatrix}
        \vphi\\
        \i{L}\\
        \i{V}
    \end{bmatrix} + \begin{bmatrix}
        \A{I} \i{s}(t)\\
        \mb{0}\\
        \v{s}(t)
    \end{bmatrix} = \mb{0}, \label{eq:md_mnaii}
\end{align}
assuming that $\phiL$ and $\qC$ do not explicitly depend on time.

We note that \eqref{eq:md_mnaii} readily extends to multiport devices\cite{tischendorf1999}. A general inductive $n$-port for example, can also be modeled by a function $\phiL(\i{L}) = \big[ \phi_1(\i{L}), \dotsc, \phi_{n-1}(\i{L}) \big]^\T$, however the Jacobian is not necessarily diagonal in this case as the component functions of $\phiL$ can each depend on all the currents $\i{L} = [i_1, \cdots, i_{n-1}]^\T$. Regarding the incidence matrices, one chooses a reference node $\varphi_0$ for the multiport device and then considers $n-1$ branches from the remaining $n-1$ nodes to the reference. Interpreting the multiport as a single node and using Kirchhoff's current law then gives $i_0 = -\sum_{k=1}^{n-1} i_k$ for the reference current, when orienting all currents to point toward the device. Following this approach, there is then no difference in treating multiports compared to one-ports\cite{tischendorf1999}.

\subsection{Index concepts}
\label{subsec:ic}
Before outlining the dissection index, we want to give a brief introduction to index concepts more generally. There are multiple definitions of the index of a DAE, along with related index concepts that each possess different strengths and weaknesses\cite{mehrmann2015}. One important aspect that unites these ideas is that they agree in key cases, e.g.~when looking at linear DAEs, and the same also holds true for the dissection index. To emphasize the practical importance of the notion of index, we take a closer look at the perturbation index. It is based on a perturbed version of the DAE \eqref{eq:int_dae} (we leave out the parameters $\mb{p}$ for conciseness)
\begin{align}
    \mb{F}(\mbh{x}', \mbh{x}, t) = \boldsymbol{\varepsilon}(t), \quad \mbh{x}(0) = \mbh{x}_0, \label{eq:ic_pdae}
\end{align}
where $\varepsilon$ is a sufficiently smooth perturbation, such that the required derivatives exist.
\begin{definition}
    \label{def:ic_pi}
    Let $\mb{x}$ be a solution of the unperturbed DAE, then the DAE is said to have perturbation index $\nu \in \mathbb{N}$, if $\nu$ is the smallest natural number such that, for any sufficiently smooth solution $\mbh{x}$ of \eqref{eq:ic_pdae}, there exists $c \in \mathbb{R}$ with
    \begin{align*}
        \| \mbh{x} - \mb{x} \| \leq c \big( \| \mbh{x}_0 - \mb{x}_0 \| + \| \boldsymbol{\varepsilon} \|_\infty + \| \boldsymbol{\varepsilon}' \|_\infty + \dotsb + \| \boldsymbol{\varepsilon}^{(\nu - 1)} \|_\infty \big)
    \end{align*}
    for an appropriate norm $\| \cdot \|$ and the right hand side small enough.
\end{definition}
The idea behind this definition is to capture the impact of perturbations on the solution, as the name suggests. Usually these pertubations are assumed small, in the sense that $\| \boldsymbol{\varepsilon} \|_\infty \ll 1$, but fast changing, such that $\| \boldsymbol{\varepsilon}^{(\nu)} \|_\infty$ may grow very quickly in $\nu$.

\paragraph{Example}
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=0.35\textwidth]{vcr}
    \end{center}
    \caption{Small example circuit for illustrating the perturbation index.}
    \label{fig:ic_vcr}
\end{figure}
To illustrate the perturbation index, and also to hint at its relevance for circuit simulation, we consider the small example given in \autoref{fig:ic_vcr}. Applying MNA to the circuit and introducing a perturbation yields the following DAE
\begin{subequations}
    \label{eq:ic_vcr}
    \begin{align}
        C \ddt \hat{\varphi}_1 + \frac{1}{R} \hat{\varphi}_1 + \hat{i}_\mr{V} &= \varepsilon_1(t) \label{eq:ic_vcra}\\
        -\hat{\varphi}_1 + v_\mr{s}(t) &= \varepsilon_2(t), \label{eq:ic_vcrb}
    \end{align}
\end{subequations}
where $\mbh{x} = [\hat{\varphi}_1, \hat{i}_\mr{V}]^\T$. Using \eqref{eq:ic_vcrb} we find
\begin{align}
    \hat{\varphi}_1 = v_\mr{s}(t) - \varepsilon_2(t). \label{eq:ic_vcrphi}
\end{align}
Inserting \eqref{eq:ic_vcrphi} into \eqref{eq:ic_vcra} and rearranging then gives
\begin{align*}
    \hat{i}_\mr{V} &= \varepsilon_1(t) - C \ddt \big( v_\mr{s}(t) - \varepsilon_2(t) \big) - \frac{1}{R} \big( v_\mr{s}(t) - \varepsilon_2(t) \big).
\end{align*}
Noting that the solution $\mb{x}$ to the unperturbed problem follows directly from the perturbed solution by setting $\boldsymbol{\varepsilon} = \mb{0}$, we obtain
\begin{align}
    \| \mbh{x} - \mb{x} \| = \begin{Vmatrix}
        -\varepsilon_2(t)\\
        \varepsilon_1(t) + C \ddt \varepsilon_2(t) + \frac{1}{R} \varepsilon_2(t)
    \end{Vmatrix}, \label{eq:ic_vcrpi}
\end{align}
so the unperturbed system corresponding to \eqref{eq:ic_vcr} has perturbation index $\nu = 2$ as \eqref{eq:ic_vcrpi} depends on the first derivative of $\varepsilon_2$.

@@

We note that all basis functions of the second step are also constant for this example, as was the case for the first step. As hinted at in \autoref{sec:di}, it is possible to find a purely topological decoupling based on the dissection index\cite{jansen2014}, which agrees with the intuition given by the index result from \autoref{th:ic_imna}. When considering large circuits, this topological decoupling along with its topological basis functions is to be preferred over other basis function choices, as it avoids the numerical computation of the basis functions, which becomes prohibitively expensive for large systems. We also note that a similar topological decoupling result exists for circuits including controlled sources, however only for the first step of the decoupling, as it is framed in the context of semi-explicit methods for which one only requires a DAE in semi-explicit form\cite{jansen2014}. An extension of this result to circuits of higher index is within the scope of further research.

@@

\paragraph{Index one case}
In order to recover the algebraic variables at time $t$ in the index one case, we only need to solve \eqref{eq:ioc_iob} for $\mbb{x}(t)$ using the learned $\mbt{x}(t)$.

\paragraph{Index two case}
In the index two case, we start by solving \eqref{eq:itc_xtp} for $\mbt{x}_\mb{P}(t)$ using the learned $\mbt{x}_\mb{Q}(t)$. Since we also require the derivative $\ddt \mbt{x}_\mb{P}(t)$ in \eqref{eq:itc_itiiib}, we consider a small time increment $\Delta t$ and approximate the derivative using a backward difference
\begin{align*}
    \ddt \mbt{x}_\mb{P}(t) \approx \frac{\mbt{x}_\mb{P}(t + \Delta t) - \mbt{x}_\mb{P}(t)}{\Delta t}.
\end{align*}
We note that this only reflects our implementation; in principle any finite difference (or similar) approximation is possible. Finally, we determine $\mbb{x}_\mb{P}(t)$ and $\mbb{x}_\mb{Q}(t)$ by jointly solving \eqref{eq:itc_xbp} and \eqref{eq:itc_itiiib} using the learned $\mbt{x}_\mb{Q}(t)$, $\mbt{x}_\mb{P}(t)$ and the approximation of $\ddt \mbt{x}_\mb{P}(t)$.

@@

There is yet another, maybe less expected, benefit that might occur. Looking at the decoupled system from \eqref{eq:itc_do2} we find that the resistance parameter $R$ and the inductance parameter $L$ only appear in the algebraic equation. In terms of our original goal of speeding up design optimization or uncertainty quantification, where a lot of solutions for varying parameter values are required, this means the following: instead of having to solve the full system for a given combination of $R$ and $L$, we can instead simply solve the algebraic equation to obtain the full solution. While the algebraic equation might be more complicated than the simple linear relations of \eqref{eq:itc_do1} and \eqref{eq:itc_do2} in general, solving it is almost certainly much faster than having to integrate the entire system in time. This becomes an even bigger advantage when knowledge about the solution is only required for specific points in time, since the algebraic equation may be solved pointwise. As of now, we have no easy way to automatically determine which parameters appear in the ODE. But when combined with a sequential learning strategy, such as the one outlined in \autoref{sec:ne}, there may still be computational savings due to the learning method requiring less samples for the parameters not appearing in the ODE.

@@

\paragraph{First example circuit}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do1_solution_phi} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{do1_solution_i_L}
    \end{center}
    \caption{Solution of the first example circuit for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$.}
    \label{fig:ne_do1_solution}
\end{figure}
We again consider the example of \autoref{fig:itc_do1}, where we choose $v_\mr{s}(t) = \sin(600 \pi t)\, \si{\volt}$, $R = \SI{500}{\ohm}$ and the diode is modeled by $g_\mr{D}(\varphi_3) = 10^{-14} \big( e^{(\varphi_3 / \SI{26}{\mV})} - 1 \big)\, \si{\siemens}$. The remaining parameter values are chosen according to the sequential sample selection strategy with $\SI{1}{\milli\henry} \leq L \leq \SI{3}{\milli\henry}$ and $\SI{100}{\nano\farad} \leq C \leq \SI{300}{\nano\farad}$, i.e.~$\mb{p} = [L, C]^\T$ in the context of \autoref{sec:int}. The starting points for the strategy are given by all combinations of the boundary values for $L$ and $C$ together with $t = \SI{0}{\ms}$ and $t = \SI{10}{\ms}$ once for each combination. We note that the choice to consider $L$ and $C$, and not for example $R$, as parameters here is arbitrary and only serves to illustrate the approach. One could still follow the same approach when considering e.g.~the initial conditions, or the diode model, as being parameterized. Recalling \eqref{eq:itc_do1}, we see that $\varphi_3$ and $i_\mr{L}$ have to be learned as the differential variables. In addition to these two, we also consider the algebraic variable $\varphi_2$ and we then compare the accuracy of learning $\varphi_2$ directly to recovering it from the two differential variables. \autoref{fig:ne_do1_solution} shows the solution for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$, a combination which is not part of the training data.
\begin{figure}[b]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do1_convergence}
    \end{center}
    \caption{Convergence of the relative prediction error $e$, for the variables shown in \autoref{fig:ne_do1_solution}. The final accuracies correspond to approximately 180 different combinations of $L$ and $C$ for $\varphi_3$, 39 combinations for $\varphi_2$ and 37 for $i_\mr{L}$.}
    \label{fig:ne_do1_convergence}
\end{figure}

@@

At this point we also return to the discussion from \autoref{sec:ial} about parameters only appearing in the algebraic equation. During the learning process, the sample selection strategy requested 21 unique values for $C$, all of which required full simulations according to \eqref{eq:itc_do2}. The 7 unique values for $L$, that were requested for learning $\varphi_2$, did not require full simulations however, but rather could be reconstructed from \eqref{eq:itc_do2}. This results from the fact that the ODE in \eqref{eq:itc_do2} does not depend on $L$, such that the solution of $\varphi_3$ also does not depend on that parameter. The same goes for the algebraic variable $i_\mr{L}$, thus $\varphi_2$ and $\varphi_1$ can be reconstructed only based on the knowledge of $\varphi_3$. We again emphasize that time is also included as a parameter within the sample selection strategy, such that the reconstructions of the algebraic variables only need to be evaluated at the particular points in time that are requested by the strategy, rather than at all time points of the solution.
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{do2_error_phi_3} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{do2_error_phi_2}
    \end{center}
    \caption{Differences between the mean prediction and simulation for $\varphi_3$ (left) and between the mean prediction, reconstruction and simulation for $\varphi_2$ (right). The predictions are again made for $L = \SI{1.7}{\milli\henry}$ and $C = \SI{220}{\nano\farad}$ and correspond to the smallest relative errors from \autoref{fig:ne_do2_solution_convergence}.}
    \label{fig:ne_do2_error}
\end{figure}

@@

\begin{figure}[b]
    \begin{center}
        \includegraphics[width=.4\textwidth]{fwr2_solution_v34} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{fwr2_solution_i_L_1}
    \end{center}
    \caption{Solutions of the differential variables of the full wave rectifier circuit for $T_1 = 65\, \si{\celsius}$ and $T_2 = 85\, \si{\celsius}$.}
    \label{fig:ne_fwr2_solution_d}
\end{figure}
As a third and larger example we consider the rectifier circuit from \autoref{fig:ne_fwr2}. Aside from requiring the full index two implementation from \autoref{sec:ial}, this example also showcases another potential application in the context of electrothermal simulations. Electrothermal simulations are used to investigate the thermal behavior of a circuit, by coupling the power that is dissipated in the circuit to a set of equations describing the temperature distribution in the circuit, and by using the temperatures of some components as parameters for certain device functions. In our case, we again model the diodes as nonlinear resistors, however this time the model also includes a temperature dependence\cite{tietze2008}
\begin{align*}
    g_\mr{D}(v_\mr{D}, T_\mr{D}) = I_\mr{s}(T_\mr{D}) \frac{qk}{T_\mr{D}} e^{(qk/T_\mr{D}) v_\mr{D}},
\end{align*}
where $v_\mr{D}$ is the voltage across the diode in forward direction, $T_\mr{D}$ is the temperature of the diode, $q$ is the charge of an electron, $k$ the Boltzmann constant and $I_\mr{s}(T_\mr{D})$ the temperature dependent reverse saturation current\cite{tietze2008}. The current source is given by $i_\mr{s}(t) = 0.1 \cos(100 \pi t)\, \si{\ampere}$ and the capacitor and resistor are linear with $C = 1\, \si{\milli\farad}$, $R = 50\, \si{\ohm}$. The circuit also contains a transformer $\mr{T}_{12}$ modeled as a nonlinear inductive multiport with
\begin{align*}
    \mb{L}(\i{L}) = \begin{bmatrix}
        L(i_{\mr{L}_1}) & \sqrt{0.09 L(i_{\mr{L}_1}) L(i_{\mr{L}_2})}\\
        \sqrt{0.09 L(i_{\mr{L}_1}) L(i_{\mr{L}_2})} & 0.1 L(i_{\mr{L}_2})
    \end{bmatrix},
\end{align*}
where $L(i_\mr{L})$ is a fourth order polynomial in $i_\mr{L}$\cite{lullo2017}. We also note that rectifiers play an important role in many power electronics applications; they are key when converting high AC to lower DC voltages. In the example we still use a current source, since this leads to an index two circuit in this case, compare \autoref{th:ic_imna}, which requires the full index two approach from \autoref{sec:ial}.
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=.4\textwidth]{fwr2_solution_phi} \hspace{1.5cm} \includegraphics[width=.4\textwidth]{fwr2_solution_i_L_2}
    \end{center}
    \caption{Solutions of the algebraic variables of the full wave rectifier circuit for $T_1 = 65\, \si{\celsius}$ and $T_2 = 85\, \si{\celsius}$.}
    \label{fig:ne_fwr2_solution_a}
\end{figure}
